Machine learning, physics and equilibrium

### 2019-05-14 distill_draft_notes/Draft/Draft.tex
html file merged

\begin{document}
\title{Physics of Equilibrium}
\author{Patrick, Juan-Miguel, Arthur, Peter, Nathan}
\maketitle

@@

Suppose we have a system which can exist in many different possible configurations $\{\vec{x}_j\}_{j=1}^J$. We assume a discrete set of configurations, but similar arguments can be made for a continuous set. We have some partial information about the system, specifically the expectation values of the random variables $\{r_k(\vec{x})\}_{k=1}^K$ (these can be constraints forced upon the system by us, or actually observed obtained via measurement). That is, we have the constraints 
\begin{equation}
 \langle r_k \rangle_{x} = a_k,
\end{equation}

@@

\textcolor{red}{Notation is not very clear yet. We probably should change it to something like: $\sum_i p_i r_{ij} = a_j$ and then follows }

@@

The Boltzmann distribution is known from physics and it describes the statistics of physical configurations of a system that is in an equilibrium at some temperature in this case the probability of a configuration $\vec{x}$ is given as $p(\vec{x}) = \frac{1}{Z} \exp{- \frac{1}{T} E(\vec{x})}$. Where $E(\vec{x})$ is a function that connects the configuration $\vec{x}$ with the real line, which is also called the energy of the system. Before we did not explicitely write the temperature because we either set it to $T=1$ or it is already absorbed in the parameters $\lambda_k$ of the model. But it is important that we are looking at systems with non-zero temperature because for zero temperature a physical system will equilibrate to its state of minimum energy. Allowing a non zero temperature automatically leads to states that are a thermal distribution which means we would observe many possible configurations that are not only of minimum energy. In the extreme case of very high temperature all the configurations even get equally likely.

@@

where $\alpha(\theta)$ is a vector valued function and $A(\theta)$ is a scalar which for a Boltzmann distribution is simply $A(\theta) = \log(1/Z)$ \cite{li_learning_2013}. Therefore in general we are not restricted to model the distribution of our data with the energy function of a classical Ising model. The choice of the constraints will determine the physical model.

@@

@@

\subsubsection{Sampling in physics}

I physics low energies are preferred over high energies. 
And therefore if we initialize a system randomly it will eventually equilibrate to a low energy configuration or if we are at $T=0$ to a local minimum of the energy.
If we imagine to have a spin model implemented as a physical device and we can initialise the spins in a random way, this spin configuration would automatically go to a configuration of lower energy.
Therefore in a real physical system, sampling from a system is only a matter of initialising it and waiting until it is equilibrate.

@@

In ML we don't have an equilibration process that drives our system into a low energy configuration. But there are several methods how we can imitate this behaviour.
To sample from a Hopfield network one takes a random input vector $\vec{v}$ and updates each node after another and then determines the update of a node via the update rule $x_i = 1,~ if~\sum_j W_{ij} x_j \geq h_i$. The outcome in this case is deterministic and one can stop this process after the nodes have been iterated through two times without any update.
For non-zero temperature models, the updates work similarly, just that the update is not deterministic, but there is a probability assigned to every node $x_i$ according to its input strength $\sum_j W_{ij} x_j + h_i$. One can imagine that

\subsubsection{Sampling from a energy based model}

Random init
Find closest local minimum
Hopfield as zero temp example (strange update rule)

@@

At the very beginning of this section we need to go one step back and briefly discuss the Hopfield network. The Hopfield network is a so called associative memory network, which does not learn patterns and generalise them to new ones, it only memories the training data and can recreate it after memorising them if one only feeds part of the data in the network. The Hopfield network has exact the same energy function $E(\vec {\sigma}) = -\sum_{i,j} w_{i,j} \sigma_i \sigma_j + \sum_i h_i \sigma_i$, but the nodes are not activated probabilistically, they follow the deterministic update rule

\begin{equation}
	\sigma_{i}\leftarrow \left\{{\begin{array}{ll}+1~{\mbox{if }}\sum _{{j}}{w_{{ij}}\sigma_{j}}\geq h _{i},\\-1~{\mbox{otherwise,}}\end{array}}\right.
\end{equation}

which can be interpreted as a Ising model at zero temperature, because this way a random input configuration $\vec{\sigma}$ will always converge to the next local minimum. No fluctuations because of temperature will occur. 
The "training" of a Hopfield network is also deterministic and we update the weight matrix simply with the outer product of the $n$ input vectors $\{ \vec{\sigma}^{\alpha}\}_i^n$ that have to be memorised. $W = \sum_{\alpha}^n( (\vec{\sigma}^{\alpha})^T \vec{\sigma}^{\alpha} - \mathbb{1}) $, which can be written as an update rule $w_{i,j} \leftarrow \sum_{\alpha}^n \sigma_i^{\alpha} \sigma_j^{\alpha}, \forall~i \neq j$
%= \sum_i^n \vec{\sigma}_i^T \vec{\sigma}_i - n \mathbb{1}$
, the subtraction of the identity matrix takes care that the nodes are not connected with themselves.
This weight matrix leads to a energy landscape where each training vector is exactly a local minimum. And if we feed a slight variation of one of the input vectors into the network, the update rule will make them converge to the configuration that is associated with the closest energy minimum. 
%After memorising the input vectors an arbitrary vector $\vec{\sigma}$ can be fed through the network and $W \vec{\sigma} =  \vec{\sigma}  \vec{\sigma}_1^T \vec{\sigma}_1 + \vec{\sigma}  \vec{\sigma}_2^T \vec{\sigma}_2 + \dots -  \vec{\sigma} n \mathbb{1} = $

@@

@@

If we now go to a bit more difficult target distribution one will realize that it cannot be approximated with the model at hand. So far we only used a Ising system to model data, where every spin represents a data point. A much more expressive model is the Boltzmann machine, where we start to distinguish between visible and hidden spin variables. On first glance there is no difference between the two models except the name of the nodes. But the crucial point is that only the visible units are used to represent data. The hidden units are only there to increase the space of possible configurations and they will be averaged out in the end. This way we can increase the expressivity of the model. In Figure (such and such) one can try again to model the harder distribution but now with hidden nodes. You will see that it is now possible to approximate more complex target distributions.

\subsubsection{Unlearning}

@@

>>>>

\begin{document}
\title{Machine learning, physics and equilibrium}
\author{Patrick, Juan-Miguel, Arthur, Peter, Nathan}
\maketitle

@@

Suppose we have a system which can exist in many different possible configurations $\{\vec{x}_j\}_{j=1}^J$. We assume a discrete set of configurations, but similar arguments can be made for a continuous set. We have some partial information about the system, specifically the expectation values of the random variables $\{r_k(\vec{x})\}_{k=1}^K$ (these can be constraints forced upon the system by us, or actually observed via measurement). That is, we have the constraints 
\begin{equation}
 \langle r_k \rangle_{x} = a_k,
\end{equation}

@@

This probability distribution is known under the name Boltzmann distribution.

@@

The Boltzmann distribution is known from physics and it describes the statistics of physical configurations of a system that is in an equilibrium at some temperature. In this case the probability of a configuration $\vec{x}$ is given as $p(\vec{x}) = \frac{1}{Z} \exp{- \frac{1}{T} E(\vec{x})}$. Where $E(\vec{x})$ is a function that connects the configuration $\vec{x}$ with the real line, which is also called the energy of the system. Before, we did not explicitely write the temperature because we either set it to $T=1$ or it is already absorbed in the parameters $\lambda_k$ of the model. But it is important that we are looking at systems with non-zero temperature because for zero temperature a physical system will equilibrate to its state of minimum energy. Allowing a non zero temperature automatically leads to states that are a thermal distribution which means we would observe many possible configurations that are not only of minimum energy, but at low energy. In the extreme case of very high temperature all the configurations even get equally likely.

@@

where $\alpha(\theta)$ is a vector valued function and $A(\theta)$ is a scalar which for a Boltzmann distribution is simply a normalization factor that we call the partition function $A(\theta) = \log(1/Z)$ \cite{li_learning_2013}. Therefore in general we are not restricted to model the distribution of our data with the energy function of a classical Ising model. The choice of the constraints will determine the physical model.

@@

\subsubsection{Architecture}

\textbf{We can make this section also as a figure. With 3 collumns, showing the models and the text below. But I think it is a good idea }

So far we have not restricted ourselves to a certain kind of model. But for practical purposes some models have been proven to be more useful than others and the most common model is based on the energy function of the Ising spin model. The first model we introduce is the Hopfield network, which is simply an Ising model of $N$ spins at zero temperature, where the dimension of the data vector is equal to the number of available spins. This kind of model does not really learn configurations, it just memorizes them.

The Boltzmann machine is in two aspect different from the Hopfield network. First it is not at zero temperature anymore, therefore we allow thermal distributions of configurations and not only local minima of the energy. And second the $N$ spins of the model are separate into $v$ visible units and $h$ hidden units. The dimension of the input data is equal to the visible units $v$.

And finally the Restricted Boltzmann machine is the model that is most commonly used. In terms of learning and expressivity of this model there is no advantage of the restricted compared to the fully-connected BM. We will see later in the training section that there is a mayor advantage of the RBM over the BM in terms of computability of the gradients.

@@

\subsubsection{Sampling in physics}
We would like to think of energy based models the following way. Imagine you are an experimentor and you have an implementation of a spin system in front of you. You can initialize the spins and you can manipulate the couplings $J_{i,j}$ and the local fields $h_i$. The only thing you can do with such a machine is to initialize the spins in some way and see to what configuration they will equilibrate.
I physics low energies are preferred over high energies. 
And therefore if we initialize a system randomly it will eventually equilibrate to a low energy configuration or if we are at $T=0$ to a local minimum of the energy.
Therefore in a real physical system, sampling from a system is only a matter of initialising it and waiting until it is equilibrate.

@@

In ML we don't have an equilibration process that drives our system into a low energy configuration. But there are several methods how we can imitate this behaviour.
To sample from a Hopfield network one takes a random input vector $\vec{v}$ and updates each node after another and then determines the update of a node via the update rule $x_i = 1,~ if~\sum_j W_{ij} x_j \geq h_i$. The outcome in this case is deterministic and one can stop this process after the nodes have been iterated through two times without any update.
For non-zero temperature models, the updates work similarly, just that the update is not deterministic, but there is a probability assigned to every node $x_i$ according to its input strength $\sum_j W_{ij} x_j + h_i$. And according to this probability we update the spins.

@@

At the very beginning of this section we need to go one step back and briefly discuss the Hopfield network \cite{Hopfield1982NeuralNA}. The Hopfield network is a so called associative memory network, which does not learn patterns and generalise them to new ones, it only memories the training data and can recreate it after memorising them if one only feeds part of the data in the network. The Hopfield network has exact the same energy function $E(\vec {\sigma}) = -\sum_{i,j} w_{i,j} \sigma_i \sigma_j + \sum_i h_i \sigma_i$, but the nodes are not activated probabilistically, they follow the deterministic update rule

\begin{equation}
\label{Eq:Hopfield update}
	\sigma_{i}\leftarrow \left\{{\begin{array}{ll}+1~{\mbox{if }}\sum _{{j}}{w_{{ij}}\sigma_{j}}\geq h _{i},\\-1~{\mbox{otherwise,}}\end{array}}\right.
\end{equation}

which can be interpreted as a Ising model at zero temperature, because this way a random input configuration $\vec{\sigma}$ will always converge to the next local minimum. No fluctuations because of temperature will occur. 
The "training" of a Hopfield network is also deterministic and we update the weight matrix simply with the outer product of the $n$ input vectors $\{ \vec{\sigma}^{\alpha}\}_{\alpha}^n$ that have to be memorised. $W = \sum_{\alpha}^n( (\vec{\sigma}^{\alpha})^T \vec{\sigma}^{\alpha} - \mathbb{1}) $, which can be written as an update rule $w_{i,j} \leftarrow \sum_{\alpha}^n \sigma_i^{\alpha} \sigma_j^{\alpha}, \forall~i \neq j$
%= \sum_i^n \vec{\sigma}_i^T \vec{\sigma}_i - n \mathbb{1}$
, the subtraction of the identity matrix takes care that the nodes are not connected with themselves. The average over all the data instances $\alpha$ is also often written as $\langle \sigma_i \sigma_j \rangle_{\text{Data}}$.
This weight matrix leads to a energy landscape where each training vector is exactly a local minimum. And if we feed a slight variation of one of the input vectors into the network, the update rule will make them converge to the configuration that is associated with the closest energy minimum. 
The memory capacity of Hopfield networks is very limited and the more data we want to memorize, the higher is the chance to get so called spurious minima. These are local energy minima which minimize the energy for configurations that are not part of the training and therefore will lead to wrong memories. 
%After memorising the input vectors an arbitrary vector $\vec{\sigma}$ can be fed through the network and $W \vec{\sigma} =  \vec{\sigma}  \vec{\sigma}_1^T \vec{\sigma}_1 + \vec{\sigma}  \vec{\sigma}_2^T \vec{\sigma}_2 + \dots -  \vec{\sigma} n \mathbb{1} = $

@@

It is worth mentioning here that in \cite{hopfield1983unlearning} Hopfield studied the effect of "unlearning" on the performance of Hopfield networks. They found that it can help to get rid of supurous minima and therefore has a stabilizing effect on the memory of such a network. 
For unlearning we expand the update rule $w_{i,j} \leftarrow \langle \sigma_i \sigma_j \rangle_{\text{Data}}$ to  $w_{i,j} \leftarrow \langle \sigma_i \sigma_j \rangle_{\text{Data}} - \epsilon \langle \sigma_i' \sigma_j' \rangle$. To obtain $\vec{\sigma}'$, the network is initialized in a random configuration $\vec{\sigma}$ and the nodes are updated according to the update rule Eq. (\ref{Eq:Hopfield update}) until the network equilibrates to $\vec{\sigma}'$. Therefore the configuration $\vec{\sigma}'$ is a minimum of the energy landscape and the term $- \epsilon \langle \sigma_i' \sigma_j' \rangle$ increases the energy of this minimum by a small factor $\epsilon<1$. This plays against the first term of the update rule $\langle \sigma_i \sigma_j \rangle_{\text{Data}}$ which minimizes the energies of the data inputs. But in the end the interplay of these two terms decreses the occurence of spurous minima.

@@

If we now go to a bit more difficult target distribution one will realize that it cannot be approximated with the model at hand. So far we only used a Ising system to model data, where every spin represents a data point. A much more expressive model is the Boltzmann machine, where we start to distinguish between visible and hidden spin variables. On first glance there is no difference between the two models except the name of the nodes. But the crucial point is that only the visible units are used to represent data. The hidden units are only there to increase the space of possible configurations and they will be averaged out in the end. This way we can increase the expressivity of the model. \textbf{hidden nodes as mediator between visible units for higher order interactions} In Figure (such and such) one can try again to model the harder distribution but now with hidden nodes. You will see that it is now possible to approximate more complex target distributions.

@@

\bibitem{Hopfield1982NeuralNA}
J.~J.~Hopfield, {\it Proceedings of the National Academy of Sciences of the United States of America}, {\bf 79 (8)}, 2554-8 (1982)

\bibitem{hopfield1983unlearning}
J.~J.~Hopfield, D.~I.~Feinstein, R.~G.~Palmer, {\it Nature}, {304}, 158, (1983)
